Supplement to Initial estuarine response to the nutrient dense Piney Point wastewater discharge into Tampa Bay, Florida

### 2021-10-21 manu-draft.Rmd
results revision done with updated figs, ord plot to supp

## Red tide and fish kill reports

```{r}
fishdat <- read.csv(here('data-raw/FishKillResultReport.csv')) %>% 
  select(
    date = textBox6, 
    city = cOUNTYDataTextBox, 
    waterbody = lOCATIONDataTextBox,
    species = textBox18
  ) %>% 
  mutate(
    date = mdy(date),
    yr = year(date),
    week = floor_date(date, unit = 'week'), 
    week = factor(format(week, '%b %d')), 
    week = factor(week, levels = as.character(unique(week))), 
    city = gsub('\\s+$', '', city),
    city = gsub('^St\\s', 'St. ', city)
  ) %>% 
  filter(city %in% c('Tampa', 'St. Petersburg')) %>% 
  filter(date < maxdt)
fishann <- fishdat %>% 
  group_by(yr, city) %>% 
  summarise(cnt = n(), .groups = 'drop') %>% 
  group_by(city) %>% 
  summarise(cnt = sum(cnt), .groups = 'drop') %>% 
  mutate(per = round(100 * cnt / sum(cnt), 1))
annstpper <- fishann %>% 
  filter(city == 'St. Petersburg') %>% 
  pull(per)
anntmpper <- fishann %>% 
  filter(city == 'Tampa') %>% 
  pull(per)
fishwk <- fishdat %>% 
  filter(yr == 2021) %>% 
  group_by(week, city) %>% 
  summarise(cnt = n(), .groups = 'drop') %>% 
  group_by(city) %>% 
  summarise(cnt = sum(cnt), .groups = 'drop') %>% 
  mutate(per = round(100 * cnt / sum(cnt), 1))
wkstpcnt <- fishwk %>% 
  filter(city == 'St. Petersburg') %>% 
  pull(cnt)
wktmpcnt <- fishwk %>% 
  filter(city == 'Tampa') %>% 
  pull(cnt)
```

The increase in *K. brevis* from April to July to bloom concentrations exceeding 10k cells/L was an anomaly in 2021 that is not regularly observed in Tampa Bay.  The historical record from 1995 to present in Figure \@ref(fig:redtide)a shows the range of cell concentrations sampled in middle and lower Tampa Bay, with only a handful of years having median cell concentrations greater than 10,000 cells/L, notably 2001 and 2012, although several years had concentrations above the median that were at bloom levels. Median cell concentrations for most years were well below 1,000 cells/L.  The highest concentration of 17.6 million cells/L was observed in 2021, whereas only 2006 had a maximum observed concentration above 10 million cells/L [@Flaherty11].  Seasonally in 2021, bloom concentrations were not observed until the week of May 23, with concentrations peaking by the week of July 4th, after which concentrations declined (Figure \@ref(fig:redtide)b).

Fish kill reports attributed to red tide at the cities of Tampa and Saint Petersburg, FL somewhat tracked the annual *K. brevis* cell concentrations, with a notable exception in 2021 which had the most reports among all years (Figure \@ref(fig:redtide)c). Historically, more fish kills were reported in Saint Petersburg (`r annstpper`%) as compared to Tampa (`r anntmpper`%), where the former is closer to the mouth of Tampa Bay.  In 2021, `r wkstpcnt` reports were made in Saint Petersburg and `r wktmpcnt` in Tampa (Figure \@ref(fig:redtide)d).  The combined weekly reports in 2021 for Tampa and Saint Petersburg peaked the week of July 4th, the same week as the peak of *K. brevis* cell concentrations (Figure \@ref(fig:redtide)b).  Notably, the first to last week of fish kill reports covered only one and half months, whereas red tide in the bay was observed for nearly four months.  Increased reports in early July coincided with a shift in winds from Tropical Storm Elsa which moved dead fish closer to populated nearshore areas, as noted earlier.  As a result, Pinellas County and the city of St. Petersburg removed over 1800 tons of dead fish near public and private shoreline areas (K. Hammer Levy, Pinellas County, pers. comm. Aug. 2021).

@@

```{r redtide, fig.cap = '*Karenia brevis* concentrations (100k cells/L) (a) by year and (b) by week in 2021, (c) fish kill reports in 2021, (d) cumulative precipitation in 2021 compared to past years, (e) cumulative inflow in 2021 compared to past years, and (f) wind rose plots for 2021 with notable breaks before/after discharge and tropical storm Elsa. *K. brevis* cell counts are from NOAA Harmful Algal BloomS Observing System (HABSOS, https://www.ncei.noaa.gov/maps/habsos), Fish kill reports are from Florida Fish and Wildlife Conservation Commission, Fish and Wildlife Research Insitute Fish Kill Database, attributed to *K. brevis* (https://public.myfwc.com/FWRI/FishKillReport/), precipitation data are from Tampa International Airport, inflow estimates are for all of Tampa Bay from USGS gaged sites, and wind rose plots are from St. Petersburg, Florida.  Wind roses show relative counts of six minute observations in directional (30 degree bins, north is vertical) and speed (m/s) categories.'}
knitr::include_graphics('figs/redtide.jpeg')
```

```{r nutrientflow, fig.cap = 'Weekly summarized observations (medians, 2.5th to 97.5th percentiles) across all sampled locations for (a) total nitrogen concentrations, (b) chlorophyll-a concentrations, (c) diatom cell concentrations, (d) *Karenia brevis* cell concentrations, and (e) cyanobacteria macroalgal abundances.  Values are summarized for all samples within each week.  The values suggest nutrient cycling between water column phytoplankton in the initial April diatom bloom, then to cyanobacteria macroalgae in May to June, and then to *K. brevis* peaking in early July. Quantitative cell counts for diatoms are missing for several weeks, but see supplementary figures for frequency occurence estimates across all dates.  Diatom concentrations are based on combined cell counts from *Asterionellopsis sp.* and *Skeletonema sp*.'}
knitr::include_graphics('figs/nutrientflow.jpeg')
```

```{r nutrientfloword, fig.cap = 'Ordination results comparing weekly summarized observations across all sampled locations for total nitrogen concentrations, chlorophyll-a concentrations, diatom cell concentrations, *Karenia brevis* cell concentrations, and cyanobacteria macroalgal abundances.  Ordination results are from non-metric multi-dimensional scaling performed on the 97.5th percentile values of observations in each week for each parameter.  Observations are grouped by month periods based on phytoplankton (diatoms or *K. brevis*) or macroalgal dominance.'}
knitr::include_graphics('figs/nutrientfloword.jpeg')
```

>>>>

## Red tide impacts

```{r}
fishdat <- read.csv(here('data-raw/FishKillResultReport.csv')) %>% 
  select(
    date = textBox6, 
    city = cOUNTYDataTextBox, 
    waterbody = lOCATIONDataTextBox,
    species = textBox18
  ) %>% 
  mutate(
    date = mdy(date),
    yr = year(date),
    week = floor_date(date, unit = 'week'), 
    week = factor(format(week, '%b %d')), 
    week = factor(week, levels = as.character(unique(week))), 
    city = gsub('\\s+$', '', city),
    city = gsub('^St\\s', 'St. ', city)
  ) %>% 
  filter(city %in% c('Tampa', 'St. Petersburg')) %>% 
  filter(date < maxdt)
fishwk <- fishdat %>% 
  filter(yr == 2021) %>% 
  group_by(week, city) %>% 
  summarise(cnt = n(), .groups = 'drop') %>% 
  group_by(city) %>% 
  summarise(cnt = sum(cnt), .groups = 'drop') %>% 
  mutate(per = round(100 * cnt / sum(cnt), 1))
wkstpcnt <- fishwk %>% 
  filter(city == 'St. Petersburg') %>% 
  pull(cnt)
wktmpcnt <- fishwk %>% 
  filter(city == 'Tampa') %>% 
  pull(cnt)
```

The increase in *K. brevis* from April to July to bloom concentrations exceeding 10,000 cells/L was an anomaly in 2021 that is not regularly observed in Tampa Bay.  The historical record from 1995 to present in Figure \@ref(fig:redtide)a shows the range of cell concentrations sampled in middle and lower Tampa Bay, with only a handful of years having median cell concentrations greater than 10,000 cells/L, notably 2005 and 2018. Median cell concentrations for most years were well below 1,000 cells/L.  The highest concentration of 17.6 million cells/L was observed in 2021, whereas only 2006 had a maximum observed concentration above 10 million cells/L [@Flaherty11].  Seasonally in 2021, bloom concentrations were not observed until the week of May 23, with concentrations peaking by the week of July 4th, after which concentrations declined (Figure \@ref(fig:redtide)b).  Cumulative rainfall and associated inflow from the main rivers entering Tampa Bay were below historical baseline values (1995 - 2020) in the months prior to the highest bloom concentrations (Figure \@ref(fig:redtide)c, d). This likely contributed to elevated salinity values in lower and middle Tampa Bay that created conditions favorable for *K. brevis* growth, in addition to the elevated nutrient concentrations from the Piney Point discharge.

Fish kill reports attributed to red tide at the cities of Tampa and Saint Petersburg, FL closely tracked cell concentrations (Figure \@ref(fig:redtide)e).  In total, `r wkstpcnt` reports were made in Saint Petersburg and `r wktmpcnt` in Tampa.  The combined weekly reports in 2021 for Tampa and Saint Petersburg peaked the week of July 4th, the same week as the peak of *K. brevis* cell concentrations (Figure \@ref(fig:redtide)b).  Notably, the first to last week of fish kill reports covered only one and half months, whereas red tide in the bay was observed for nearly four months. Increased fish kill reports in early July may coincided with a shift in winds from Tropical Storm Elsa (Figure \@ref(fig:redtide)f), which moved dead fish closer to populated nearshore areas, as noted earlier.  By August, Pinellas County and the city of St. Petersburg removed over 1800 tons of dead fish near public and private shoreline areas (K. Hammer Levy, Pinellas County, pers. comm. Aug. 2021).

## Nutrient cycling

The above results can be considered together to develop a concise narrative of the temporal shift of nutrient pools between different ecosystem components of the bay from April to September, starting with the influx of inorganic nitrogen from Piney Point. Total nitrogen concentrations first peaked in April (Figure \@ref(fig:nutrientflow)a), as did chlorophyll-a concentrations (Figure \@ref(fig:nutrientflow)b).  This initial peak in water quality parameters indicated a unique response of the phytoplankton community as an increase in diatoms (Figure \@ref(fig:nutrientflow)c) that can readily utilize inorganic forms of nitrogen (e.g., centric species, also *Asterionellopsis sp.* and *Skeletonema sp.*).  Water quality indicators improved slightly following the decrease in diatoms in late April, as noted by relatively lower concentrations of total nitrogen and chlorophyll.  However, macroalgal biomass increased after the initial diatom bloom and peaked in June (Figure \@ref(fig:nutrientflow)d), suggesting a shift of nutrients from phytoplankton to drift macroalgae communities. During peak macroalgae growth, total nitrogen and chlorophyll-a concentrations remained relatively low as nutrients were likely retained in macroalgae, until late June and early July when *K. brevis* concentrations peaked (Figure \@ref(fig:nutrientflow)e).  The co-occurring decline in macroalgae and increase in *K. brevis* suggests a release of nutrients from the former that stimulated growth of the latter, although residual water column nutrients were likely present (as noted by modelling efforts).  Conditions were relatively stable in August and September with relatively improved water quality conditions and no dominant algal blooms.  These distinct temporal periods were readily identified through an ordination plot (see supplement) for the observed data in Figure \@ref(fig:nutrientflow).

@@

```{r redtide, fig.cap = '*Karenia brevis* concentrations (100k cells/L) (a) by year and (b) by week in 2021, (c) cumulative precipitation in 2021 compared to past years, (d) cumulative inflow in 2021 compared to past years, (e) fish kill reports in 2021, and (f) wind rose plots for 2021 with notable breaks before/after discharge and tropical storm Elsa. *K. brevis* cell counts are from NOAA Harmful Algal BloomS Observing System (HABSOS, https://www.ncei.noaa.gov/maps/habsos), Fish kill reports are from Florida Fish and Wildlife Conservation Commission, Fish and Wildlife Research Insitute Fish Kill Database, attributed to *K. brevis* (https://public.myfwc.com/FWRI/FishKillReport/), precipitation data are from Tampa International Airport, inflow estimates are for all of Tampa Bay from USGS gaged sites, and wind rose plots are from St. Petersburg, Florida.  Wind roses show relative counts of six minute observations in directional (30 degree bins, north is vertical) and speed (m/s) categories.'}
knitr::include_graphics('figs/redtide.jpeg')
```

```{r nutrientflow, fig.cap = 'Weekly summarized observations (medians, 2.5th to 97.5th percentiles) across all sampled locations for (a) total nitrogen concentrations, (b) chlorophyll-a concentrations, (c) diatom cell concentrations, (d) cyanobacteria macroalgal abundances, and (e) *Karenia brevis* cell concentrations.  Values are summarized for all samples within each week.  The values suggest nutrient cycling between water column phytoplankton in the initial April diatom bloom, then to cyanobacteria macroalgae in May to June, and then to *K. brevis* peaking in early July. Quantitative cell counts for diatoms are missing for several weeks, but see supplementary figures for frequency occurence estimates across all dates.  Diatom concentrations are based on combined cell counts from *Asterionellopsis sp.* and *Skeletonema sp*.'}
knitr::include_graphics('figs/nutrientflow.jpeg')
```

### 2021-10-21 supplement.Rmd
results revision done with updated figs, ord plot to supp

>>>>

```{r nutrientfloword-supp, fig.cap = 'Ordination results comparing weekly summarized observations across all sampled locations for total nitrogen concentrations, chlorophyll-a concentrations, diatom cell concentrations, *Karenia brevis* cell concentrations, and cyanobacteria macroalgal abundances.  Ordination results are from non-metric multi-dimensional scaling performed on the 97.5th percentile values of observations in each week for each parameter.  Observations are grouped by month periods based on phytoplankton (diatoms or *K. brevis*) or macroalgal dominance.'}
knitr::include_graphics('figs/nutrientfloword-supp.jpeg')
```
